---
title: "Tidy Data Demo"
output: html_document
---

Tidying data: walking through [R4DS tidy data chapter](https://r4ds.had.co.nz/tidy-data.html)

## Intro

```{r}
## load libraries
library(tidyverse)

## examine tidyverse built-in datasets
## same data, four different ways. Each dataset shows the same values of four variables country, year, population, and cases, but each dataset organises the values in a different way.

table1
table2
table3
table4a
table4b
```

Do any of these look familiar? 

Having your variables in columns lets you leverage R's "vectorized" nature, meaning that you can operate across a whole vector of values. Let's see: 

```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>% 
  count(year, wt = cases) # wt is frequency weighting

# Visualise changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```

Exercises?

## Pivoting
Data is often organised to facilitate some use other than analysis. For example, data is often organised to make entry as easy as possible.

2 main things to do when you encounter untidy data: 

1. figure out what the variables and observations are. Sometimes this is easy; other times you’ll need to consult with the people who originally generated the data. 

2. resolve one of two common problems:

2a. One variable might be spread across multiple columns.

2b. One observation might be scattered across multiple rows.


### Longer
A common problem is a dataset where some of the column names are not names of variables, but values of a variable. Take table4a: the column names 1999 and 2000 represent values of the year variable, the values in the 1999 and 2000 columns represent values of the cases variable, and each row represents two observations, not one.

```{r}
table4a

tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")

tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")

left_join(tidy4a, tidy4b)

```

### Wider

Wide data has observations scattered across multiple rows. For example, take table2: an observation is a country in a year, but each observation is spread across two rows.

```{r}
table2
table2 %>%
    pivot_wider(names_from = type, values_from = count)

```

Exercises

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")
```

### Separate

```{r}
table3

table3 %>% 
  separate(rate, into = c("cases", "population"))

# convert to numeric
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)

```

### Missing Values

complete()

https://www.imachordata.com/2016/02/05/you-complete-me/


